---
title: An R Markdown document converted from "./book/9_hypothesis_tests.ipynb"
output: html_document
---

# 9. Hypothesis Testing

In this chapter, we will look at hypothesis testing in R. We will start with single sample distributions and tests, and then we will look at hypothesis tests when comparing two samples. Examples will include testing for positive correlations, performing two sample paired t-tests, and testing for equal variance among groups. The data we will use in this section comes from the Texas Health and Human Services Department and includes the reported number of induced terminations of pregnancy (ITOPs) from 2016 to 2021, stratified by both race and county. The data also contains the rate of abortions per 1000 females aged 15-49. Read the data documentation to see the full variable descriptions.

```{r}
library(tidyverse)
library(car)
library(RforHDSdata)
library(gt)
library(gtsummary)
data(tex_itop)
```

## Univariate Distributions and One Sample Tests

Let's begin by looking at a single outcome of interest - the number of induced terminations of pregnancy (referred to as ITOPs or abortions below) in 2021 per 1000 females ages 15-49 in each county. We use the number of females ages 15-49 as a proxy to scale the number of abortions by the population size, though this is not truly reflective of the number of people who can give birth in each county. 

```{r}
county_rates_2021 <- tex_itop$total_rate[tex_itop$year == 2021]
hist(county_rates_2021, breaks = 35)
```

We can see in the figure that this is a heavy-tailed distribution. Below, we find the 10 counties with highest rates and see that there are some counties with very few total abortions but that have some of the highest abortion rates. This indicates a small population. On the other hand, we also observe Harris county, which contains the city of Houston and has both a high total abortion count and a high abortion rate. 

```{r}
tex_itop %>% 
  filter(year==2021) %>% 
  slice_max(n=10, total_rate) %>%
  dplyr::select(c(county, total_itop, total_rate))
```

Some of the counties are so small that we may want to consider dropping them from our analysis. In particular, the rates in Loving County and Terrel County are so high that we might consider them to be outliers. For this one sample analysis, however, we do not remove them. If we wanted to estimate the mean abortion rate among counties $\mu$ we can do so by simply using the `mean()` function. For reference, the [CDC](https://www.cdc.gov/reproductivehealth/data_stats/abortion.htm) reported the national abortion rate to be 11.2 abortions per 1,000 women aged 15–44 years.

```{r}
mean(county_rates_2021, na.rm=TRUE)
```

Within R we can also calculate a confidence interval for this mean. Recall that a $(1-\alpha)$%  confidence interval for the mean is given by the equation below, where $\hat{\mu}$ is our sample mean, $\hat{\sigma}^2$ is the sample variance, and $n$ is the number of observations.
$$\hat{\mu} \pm z_{1-\alpha/2} \cdot \frac{\hat{\sigma}}{\sqrt{n}}$$

Below, we use this formula to calculate a 95% confidence interval for the mean abortion rate among counties:

```{r}
est_mean <- mean(county_rates_2021, na.rm=TRUE)
est_sd <- sd(county_rates_2021)
z_alpha <- dnorm(1-0.05/2)
n <- length(county_rates_2021)
c(est_mean - z_alpha*est_sd/sqrt(n), est_mean + z_alpha*est_sd/sqrt(n))
```

If we want to display this nicely, we can use the `round()` function, which allows us to specify a number of digits to be displayed, and the `paste()` function, which creates a single character string from multiple inputs. 

```{r}
lower <- round(est_mean - z_alpha*est_sd/sqrt(n),3)
upper <- round(est_mean + z_alpha*est_sd/sqrt(n),3)
paste("Confidence Interval: (", lower, ",", upper, ")")
```

Suppose that we wanted to run a hypothesis test to compare the mean to a pre-determined value. In particular, the Texas Heartbeat Act was introduced in 2021 and drastically reduced the number of eligible abortions. We could test whether there were significantly fewer abortions in 2021 compared to 2020. Our null hypothesis is that $\mu \geq 6.23$, the mean abortion rate in 2020. To run this hypothesis test, we use the `t.test()` function. For a one sample t-test, we need to specify our sample `x`, the alternative hypothesis `alternative` (default is a two-sided test), the true value of the mean `mu` (default 0), and a confidence level `conf.level` (default 0.95). Below, we run this t-test, and we can see from the result that we reject the null hypothesis and observe a statistically significant difference in the mean abortion rate between 2020 and 2021. 

```{r}
t.test(county_rates_2021, alternative = "less", mu = 6.23, conf.level=0.95)
```

One thing to consider is that the `t.test()` function assumes that the sample `x` comes from a normal distribution. The one-sample Wilcoxon signed rank test is a non-parametric alternative to the one-sample t-test that can be used to compare the median value of a sample to a theoretical value without assuming that the data is normally distributed. This test can be performed using the `wilcox.test()` function and takes in the same arguments as the `t.test()` function. Below, we can see that we again reject the null hypothesis and conclude that the mean abortion rate in 2021 was significantly lower than that of 2020.

```{r}
wilcox.test(county_rates_2021, alternative = "less", mu = 6.23, conf.level=0.95)
```

## Correlation and Covariance

We now look at two sample tests. To start, we look at the 2020 and 2021 rates by county. We pivot our data into a wider format in order to create 2020 and 2021 rate columns, and, this time, we filter out the Loving and Terrel counties to remove outliers. We then create a scatter plot of 2021 vs 2020 rates and observe a linear correlation between the two.

```{r}
county_rates <- tex_itop %>%
  dplyr::select(c(county, total_rate, year)) %>%
  filter(!(county %in% c("Terrell", "Loving")), year %in% c(2020, 2021)) %>%
  pivot_wider(names_from = year, values_from = total_rate) %>%
  na.omit() %>%
  rename("y2020"="2020", "y2021"="2021")
head(county_rates)
```

```{r}
ggplot(county_rates) + 
 geom_point(aes(x=y2020,y=y2021)) +
 labs(x="2020 ITOP Rates", y="2021 ITOP Rates")
```

We have seen before how to calculate the correlation between two columns using the `cor()` function. We can also calculate the covariance using the `cov()` function. As suspected, there is a strong correlation. The estimated covariance is around 5.19.

```{r}
cor(county_rates$y2020, county_rates$y2021)
cov(county_rates$y2020, county_rates$y2021)
```

Besides calculating the value of the correlation, we can also test whether this correlation is significantly different from zero. The function `cor.test()` tests for association between paired samples, using either Pearson's product moment correlation coefficient, Kendall's $\tau$, or Spearman's $\rho$. Similar to the `t.test()` and `wilcox.test()` functions, we can also specify the `alternative` and `conf.level` arguments. Below, we test whether there is a non-zero correlation between the 2020 and 2021 county rates using Pearson's product-moment correlation. We can see from the resulting p-value that we can reject the null hypothesis that the correlation is zero and conclude that it is instead significantly different than zero. 

```{r}
cor.test(county_rates$y2020, county_rates$y2021, method="pearson")
```

## Two Sample Tests for Continuous Variables

If we wanted to directly compare the difference between 2020 and 2021 rates, we could use a two sample test. In this case, because our samples are paired by county, we can use a two sample paired t-test. Specifically, we use a two-sided test to test the null hypothesis that the rates are equal by specifying two different vectors `x` and `y`. Note that we used the default values of `mu=0` and `alternative="two.sided"`. Additionally, we used the default value `var.equal=FALSE`, which implies that the samples may have different variances. From the results below, we reject the null hypothesis that the two county rates are equal. 

```{r}
t.test(x=county_rates$y2020, y=county_rates$y2021)
```

In the `tex_itop` dataset, each county has also been categorized by whether it was urban or rural. Suppose we want to compare the change in abortion rates from 2020 to 2021 between rural and urban counties. First, we create a variable describing the rate change between these years using the code below. We choose to use the change in rate rather than percent change to avoid infinite or undefined values.

```{r}
county_rates_urban <- tex_itop %>%
  dplyr::select(c(county, urban, total_rate, year)) %>%
  filter(total_rate < 15, year %in% c(2020, 2021)) %>%
  pivot_wider(names_from = year, values_from = total_rate) %>%
  na.omit() %>%
  rename("y2020"="2020", "y2021"="2021") %>%
  mutate(rate_change = (y2021-y2020)) 
```

We again use a two-sample two-sided t-test, but this time the data is not paired. Below, we show an alternative way to specify a t-test test below using a formula `lhs ~ rhs`, where `lhs` is a numeric vector and `rhs` is a factor variable with two levels. We must also specify the data in this case. From the R output in this case, we would fail to reject the null hypothesis and conclude that the rate changes for urban and rural counties are not significantly different. 

```{r}
t.test(rate_change~urban, data=county_rates_urban)
```

Besides a t-test, we can also use a two-sample Wilcoxon non-parametric test using the `wilcox.test()` function, which has the same arguments as the function `t.test()`. Both the `t.test()` and `wilcox.test()` can only compare two groups. When we want to compare two or more independent samples, we can use a Kruskal-Wallis rank sum test using the `kruskal.test()` function or a one-way analysis of variance (ANOVA) using the `aov()` function. Even though we only have two groups in our example, we show how to run the corresponding tests for these two groups below. For the `kruskal.test()` function, we can either specify the arguments `formula` and `data` or we can specify two vectors: `x`, a numeric vector, and `g`, a factor representing the group. For the `aov()` function, we specify the test using a formula and then specify the data. To see the p-value, we have to use the `summary()` function to print the result. Again, both tests suggest that we fail to reject the null hypothesis.

```{r}
kruskal.test(county_rates_urban$rate_change, county_rates_urban$urban)
```

```{r}
aov_res <- aov(rate_change~urban, data=county_rates_urban)
summary(aov_res)
```

We could also test the variance of a continous variable between two groups using an F test. Our null hypothesis for this test is that the variance in both groups is equal. The function `var.test()` implements an F test and has the same main arguments as the `t.test()` function: vectors `x` and `y` OR a `formula` and `data`, the alternative hypothesis `alternative`, and `conf.level`. Additionally, we can specify the hypothesized ratio of the variances through the arugment `ratio` (default value 1). Note that this function assumes that the two samples come from normally distributed populations.

```{r}
var.test(y2021 ~ urban, county_rates_urban)
```

Lastly, we can consider Levene's test to test whether group variances are equal when there are more than two groups. This test can be specified using a formula and data set, as below, or by providing two vectors `y`, a numeric vector, and `g`, a vector specifying the groups. 

```{r}
leveneTest(y2021 ~ as.factor(urban), county_rates_urban)
```

## Two Sample Tests for Categorical Variables

In the two sample tests above, we were comparing the distributions of continuous variables. We now look at comparing distributions of categorical variables. We will first categorize counties by their abortion rate in 2020 being above or below 11.2, which was the national average rate that year. We display the distribution of this variable by the urban/rural grouping using a contingency table below.

```{r}
county_rates_urban$below_nat_avg <- ifelse(county_rates_urban$y2020 > 11.2, "Above Nat Avg", "Below Nat Avg")
table(county_rates_urban$below_nat_avg, county_rates_urban$urban)
```

We can use a Fisher's exact test to test the distribution in this contingency table. The `fisher.test()` function can either take in a contingency table as a matrix or can be specified by two factor vectors `x` and `y`. Additionally, there is the option to specify the `alternative` and `conf.level` arguments. Again, we do not see a significant difference between urban and rural counties.

```{r}
fisher.test(county_rates_urban$below_nat_avg, county_rates_urban$urban)
```

An alternative test is a Pearson's Chi-Squared test, which can be used for large sample sizes. The counts of rural and  urban counties in the 'Above Nat Avg' category are very small, so we recategorize our outcome to be at or above Texas's average to avoid this complication. The `chisq.test()` function also takes in a contingency table as a matrix or can be specified by two factor vectors `x` and `y`. Another useful argument is `correct` (default is TRUE) which indicates whether to apply a continuity correction. 

```{r}
tex_mean <- mean(county_rates_urban$y2020)
county_rates_urban$below_tex_avg <- ifelse(county_rates_urban$y2020 > tex_mean, 
                                           "Above Texas Ave", "Below Texas Ave")
table(county_rates_urban$below_tex_avg, county_rates_urban$urban)
```

```{r}

chisq.test(county_rates_urban$below_tex_avg, county_rates_urban$urban)
```

## Adding Hypothesis Tests to Summary Tables

Earlier, we used the `gt` and `gtsummary` packages to create summary tables of variables. When creating a stratified table (done by add the `by` argument), we can automatically add p-values for hypothesis tests comparing across populations using the `add_p()` function. By default, the `add_p()` function uses a Kruskal-Wallis rank sum test for continuous variables (or a Wilcoxon rank sum test when the `by` variable has two levels) and uses a Chi-Squared Contingency Table Test for categorical variables (or a Fisher's Exact Test for categorical variables with any expected cell count less than five). The chosen test(s) are dispalyed as footnotes.

```{r}
tbl_summary(tex_itop, include = c(total_rate, white_rate, asian_rate, hispanic_rate, 
                                  black_rate, native_american_rate),
           by = "year", statistic = list(all_continuous() ~ "{mean} ({sd})")) %>% 
  add_p() %>%
  as_gt() %>% 
  gt:::as.tags.gt_tbl()
```

## Exercises

Use different data and ask to recreate summary table p-values

```{r}
data(pain)
```

